Long name of the paper

### 2018-10-28 anzjs2018/sections/01_introduction.tex
Edits

The need for realtime information.
Where it goes wrong: 
- bad schedule calibration;
- no accounting for realtime traffic 
- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
Literature review of historical and present methods?

% Bus arrival-time prediction is far from a novel concept: 
% ever since the first vehicle tracking technology was implemented
% in transit vehicles in the late 1980's, 
% real-time models have been implemented to provide \gls{rti} for commuters.
% The simplest form of this is, the \gls{eta}
% (usually displayed as \emph{minutes until arrival}),
% was implemented in {{region}} by \cite{Wall99analgorithm} 
% using their proposed \gls{kf} algorithm.
% In their work, they used historical data to obtain \glspl{eta}.

% Over the years, advances in computing power have enabled new
% modeling approaches,
% such as support vector machines \citep{Yu_2006},
% artificial neural networks \citep{Yu_2011},
% and more recently particle filters \citep{Hans_2015}.

Particle filter has proven useful to this kind of dynamic problem,
where the state distribution is non-gaussian, non-symmetric, 
multimodal...

Any other recent attempts at network-estimation?
There was a KF using point-to-point speed estimations?

Computational aspect - historically, models had to be simple 
(e.g., KF) so they could run in realtime.
These days, running a somewhat powerful VM on a remote server
is almost standard practice, so we can take full advantage 
and use a more computationally demanding (particle filter)
but more accurate estimation technique - paired with a more complex model.

We first give an overview of the transit network,
and how we construct it from raw GTFS shape data.
This paper assumes a very basic model,
with a focus on the implementation in a realtime setting.

>>>>

In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and in some places make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
A second is that buses are shown as on time but in fact are not running,
and therefore have not registered on the \rt system.
In this case, the scheduled times are used for ETAs and, 
when the bus doesn't show up, 
the bus disappeared from the \rt board and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as accurate as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA can change drastically.

To avoid these problems and improve the reliability of ETAs,
we first needed to develop a model to infer position and speed
of transit vehicles from \rt GPS data,
isntead of relying on delays reported only at stops.
Several modeling approaches were explored, 
such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter approach demonstrated
by \cite{Hans_2015}.
We chose to use the last of these, the particle filter,
as they are very flexible and aren't limited by assumptions
such as unimodality;
instead, they are constrained by computational complexity.

The \emph{reliablility} of arrival time estimates
alo needs to be considered.
Ideally, congestion along intermediate roads should be incorporated into the predictive model,
and one way of doing this is to use the travel time of preceeding buses along the same roads.
Several such approaches have been proposed.
\cite{Yu_2010} used the travel time of vehilces on the same \emph{routes},
which they showed to improve ETAs.
However, this only works on routes with a high frequency.
\citep{Yu_2011} extended this by looking at roads where several major rouets converged,
again showing improvements but limited in scope.
In general, however, no comprehensive network modeling approach has been proposed that used
solely GPS position data to estimate arrival times
while accounting for congestion.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in \rt.
The vehicle model is implemented using a particle filter,
so it is also necessary to demonstrate the faesibility of this as a valid real-time option.
From the particle filter estimates of vehicle trajectories,  
we estimate the travel time along roads update the road network's state.
Future work will involve using this travel time information 
in combination with historical data to develop an arrival time prediction model.

### 2018-10-28 anzjs2018/sections/02_gtfs.tex
Edits

\section{Working with GTFS data}
\label{sec:gtfs}

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who make use of it in their Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this for us is that,
provided our application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about stops, routes, trips, and shapes:
a \emph{stop} is a physical location where passengers can embark and disembark the vehicle;
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day; and
a \emph{shape} is the sequence of points defining a vehicle's path along a route
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
onboard GPS and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
which are often accessed via an API which developers can use in \rt applications.

Before we can develop our application,
we need to modify the raw GTFS information to make it possible to detect common roads.

@@

Arguably one of the most important predictors of arrival time is
the travel time along intermediate roads,
however in most applications this vital information is unavailable,
at least directly. 
Several predictive approaches using \emph{headway},
the time between consecutive trips along the same route,
have been proposed \citep{Hans_2015};
however, while this is reasonable for high frequency routes,
low frequncy routes will be unable to react quickly to changes in congestion,
and it is for these routes which reliable ETAs are arguably more important,
since the cost of missing a bus is greater.

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times.
The simplest method of identifying overlapping routes is to compare
the stop sequence: 
routes with a common sub-sequence of stops must be traveling the same route between them.
While there are several exceptions, for example express routes 
and multi-stop locations where there are high numbers of routes,
it is a viable enough simplication for the purposes of this paper.
Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is one or more bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive 
(i.e., the bus is almost there, but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection
(e.g., in a queue of traffic at traffic lights).
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the previous observation is rejected and the vehicle reverted
to its previous state.

>>>>

\section{Working with \rt transit data}
\label{sec:gtfs}

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who make use of it in their Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occuring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
onboard GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current ETA generation method.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs (in Auckland).
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes most importantly the delay between the scheduled and actual arrival times.
This delay is then propagated on to all future stops to adjust the ETA.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
Even more problematic is that, in the absense of trip updates,
the delay defaults to zero,
and unless the trip is manually cancelled,
appears on-time to passengers waiting at the stop.
If the bus is late or cancelled,
the trip disappears from the \rt board and passengers
are unable to distingush the two scenarios.

@@

Arguably the most important predictor of arrival time is
the cumulative travel time along intermediate roads.
In most applications, however, this vital information is unavailable,
at least directly,
so here we attempt to construct a transit road network of intersections
and the roads between them, 
and map each route to a sequence of road segments.

The simplest way to construct such a network is to use the stop sequence:
routes with a common sub-sequence of stops must be traveling the same route between them,
so by defining a road as the path between two consecutive stops 
we can obtain travel time estimates along that road segment.
There are several places where this doesn't quite work, for example express routes 
and multi-stop locations where there are high numbers of routes,
resulting in some overlap between road segments,
it is a viable enough simplication for the purposes of this paper.
Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.
Future work will look at using physical intersections as nodes in the network to improve it.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is one or more bus stop
or intersection between them.
It is also possible for a bus to remain stationary,
so the possible vehicle positions rapidly increases with the time between observations.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these updates are preemptive 
(i.e., the bus is almost there, but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection
(e.g., in a queue of traffic at traffic lights).
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state.
